---
title: "Cohen's d and ANOM"
author: "Dan Swart"
date: "2025-04-29"
format: html
---


```{r}
#| label: setup
#| include: false
#| echo: false


knitr::opts_chunk$set(echo = TRUE)
```

# Visualizing Distribution Overlap

```{r}
#| echo: false

# Example code for visualizing overlap
library(ggplot2)

# Simulate data with small mean difference but large overlap
group_A <- stats::rnorm(1000, mean = 100, sd = 15)
group_B <- stats::rnorm(
  1000,
  mean = 95,
  sd = 15
)

# Put in data frame
df <- base::data.frame(
  score = base::c(group_A, group_B),
  group = base::factor(base::c(base::rep("Group A", 1000), base::rep("Group B", 1000)))
)

# Calculate Cohen's d
cohens_d <- (base::mean(group_A) - base::mean(group_B)) /
  base::sqrt((stats::var(group_A) + stats::var(group_B)) / 2)

# T-test
t_result <- stats::t.test(group_A, group_B)

# Plot distributions
ggplot2::ggplot(df, ggplot2::aes(x = score, fill = group)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::geom_vline(xintercept = base::mean(group_A), color = "blue", linetype = "dashed") +
  ggplot2::geom_vline(xintercept = mean(group_B), color = "red", linetype = "dashed") +
  ggplot2::labs(
    title = "Distribution Overlap Despite Different Means",
    subtitle = paste(
      "Cohen's d =",
      base::round(cohens_d, 2),
      ", p-value =",
      base::format.pval(t_result$p.value, digits = 3)
    ),
    x = "Score",
    y = "Density"
  ) +
  ggplot2::theme_minimal()
```

# Quantifying Distribution Overlap

Calculate the overlap coefficient and probability of superiority from Cohen's d:

```{r}
#| echo: false


#| # Function to calculate overlap coefficient from Cohen's d
calculate_overlap <- function(d) {
  # Overlap coefficient formula
  ovl <- 2 * stats::pnorm(-abs(d) / 2) * 100
  base::return(ovl)
}

# Function to calculate probability of superiority from Cohen's d
calculate_superiority_prob <- function(d) {
  # Probability of superiority formula
  prob <- stats::pnorm(d / base::sqrt(2)) * 100
  base::return(prob)
}

# Example with your Cohen's d = 0.36
d_value <- 0.36
overlap_percent <- calculate_overlap(d_value)
superiority_percent <- calculate_superiority_prob(d_value)

# Output results
cat("Cohen's d =", d_value, "\n")
cat("Overlap coefficient =", round(overlap_percent, 1), "%\n")
cat("Probability of superiority =", round(superiority_percent, 1), "%\n")

# Create visualization of overlap
library(ggplot2)

# Generate data for two normal distributions
x <- base::seq(-4, 4, length.out = 1000)
group1 <- base::data.frame(x = x, y = stats::dnorm(x, mean = 0, sd = 1), group = "Group A")
group2 <- base::data.frame(
  x = x,
  y = stats::dnorm(x, mean = d_value, sd = 1),
  group = "Group B"
)
combined_data <- base::rbind(group1, group2)

# Create plot
overlap_plot <- ggplot2::ggplot(combined_data, ggplot2::aes(x = x, y = y, fill = group)) +
  ggplot2::geom_area(alpha = 0.5, position = "identity") +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "blue") +
  ggplot2::geom_vline(xintercept = d_value, linetype = "dashed", color = "red") +
  ggplot2::annotate(
    "text",
    x = -1.5,
    y = 0.1,
    label = paste0("Overlap = ", base::round(overlap_percent, 1), "%")
  ) +
  ggplot2::annotate(
    "text",
    x = -1.5,
    y = 0.15,
    label = base::paste0("Cohen's d = ", d_value)
  ) +
  ggplot2::annotate(
    "text",
    x = -1.5,
    y = 0.05,
    label = base::paste0("P(superiority) = ", base::round(superiority_percent, 1), "%")
  ) +
  ggplot2::labs(
    title = "Distribution Overlap for Cohen's d",
    subtitle = "Showing the extent of overlap between groups with different means",
    x = "Standardized Score (z-score)",
    y = "Density"
  ) +
  ggplot2::theme_minimal()

print(overlap_plot)

# To apply this to your actual groups:
# Replace with your actual groups' data
# group_A <- your_first_group_data
# group_B <- your_second_group_data

# Calculate means and SDs and pool them
# mean_A <- mean(group_A)
# mean_B <- mean(group_B)
# sd_pooled <- sqrt((var(group_A) + var(group_B))/2)

# Calculate Cohen's d
# cohens_d <- (mean_A - mean_B) / sd_pooled

# Calculate overlap and superiority probability
# overlap <- calculate_overlap(cohens_d)
# superiority <- calculate_superiority_prob(cohens_d)
```


# Simulating distributions when you only have means and SDs:

```{r}
#| echo: false


# Example: Simulating distributions from summary statistics
library(ggplot2)

# Given summary statistics
male_mean <- 78.5
male_sd <- 12.3
female_mean <- 80.2
female_sd <- 11.6
male_n <- 500 # Estimate sample size
female_n <- 500 # Estimate sample size

# Simulate individual-level data
set.seed(123)
male_scores <- stats::rnorm(male_n, mean = male_mean, sd = male_sd)
female_scores <- stats::rnorm(female_n, mean = female_mean, sd = female_sd)

# Calculate Cohen's d
cohens_d <- (female_mean - male_mean) / base::sqrt((male_sd^2 + female_sd^2) / 2)

# Calculate overlap coefficient
overlap <- calculate_overlap(cohens_d)
superiority <- calculate_superiority_prob(cohens_d)

# Create data frame for plotting
df <- base::data.frame(
  score = c(male_scores, female_scores),
  gender = base::factor(base::c(base::rep("Male", male_n), base::rep("Female", female_n)))
)

# Visualize the distributions
ggplot2::ggplot(df, ggplot2::aes(x = score, fill = gender)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::geom_vline(xintercept = male_mean, color = "blue", linetype = "dashed") +
  ggplot2::geom_vline(xintercept = female_mean, color = "red", linetype = "dashed") +
  ggplot2::annotate(
    "text",
    x = base::max(df$score) * 0.8,
    y = 0.03,
    label = base::paste0("Cohen's d = ", base::round(cohens_d, 2))
  ) +
  ggplot2::annotate(
    "text",
    x = max(df$score) * 0.8,
    y = 0.02,
    label = paste0("Overlap = ", round(overlap, 1), "%")
  ) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "Simulated Distributions Based on Summary Statistics",
    subtitle = "Showing how male and female scores might be distributed"
  )

```

# Analysis of Means (ANOM)

```{r}
#| echo: false

library(ggplot2)
library(dplyr)

# Given summary statistics
male_mean <- 78.5
male_sd <- 12.3
female_mean <- 80.2
female_sd <- 11.6
male_n <- 500 # Estimated sample size
female_n <- 500 # Estimated sample size

# Calculate overall (grand) mean
overall_n <- male_n + female_n
overall_mean <- (male_mean * male_n + female_mean * female_n) / overall_n

# Calculate pooled standard deviation
pooled_sd <- base::sqrt(
  ((male_n - 1) * male_sd^2 + (female_n - 1) * female_sd^2) /
    (male_n + female_n - 2)
)

# Calculate standard error for each group
male_se <- pooled_sd / base::sqrt(male_n)
female_se <- pooled_sd / base::sqrt(female_n)

# Calculate critical value (for 95% confidence)
# Using t-distribution with df = n1 + n2 - 2
alpha <- 0.05
critical_value <- stats::qt(1 - alpha / 2, df = male_n + female_n - 2)

# Calculate decision limits
upper_decision_limit <- overall_mean +
  critical_value * pooled_sd * base::sqrt(1 / male_n + 1 / female_n)
lower_decision_limit <- overall_mean -
  critical_value * pooled_sd * base::sqrt(1 / male_n + 1 / female_n)

# Create data frame for ANOM plot
anom_df <- base::data.frame(
  group = base::c("Male", "Female"),
  mean = base::c(male_mean, female_mean),
  n = base::c(male_n, female_n)
)

# Calculate if means are significantly different from overall mean
anom_df$significant <- (anom_df$mean > upper_decision_limit) |
  (anom_df$mean < lower_decision_limit)

# Create ANOM plot
ggplot2::ggplot(anom_df, ggplot2::aes(x = group, y = mean, color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_mean, linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_decision_limit, linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_decision_limit, linetype = "dashed") +
  ggplot2::annotate(
    "text",
    x = 1.5,
    y = overall_mean + 1,
    label = paste("Overall Mean =", base::round(overall_mean, 2))
  ) +
  ggplot2::annotate(
    "text",
    x = 1.5,
    y = upper_decision_limit + 1,
    label = "Upper Decision Limit"
  ) +
  ggplot2::annotate(
    "text",
    x = 1.5,
    y = lower_decision_limit - 1,
    label = "Lower Decision Limit"
  ) +
  ggplot2::scale_color_manual(values = c("black", "red")) +
  ggplot2::labs(
    title = "Analysis of Means (ANOM)",
    subtitle = "Comparing Male and Female Score Means to Overall Mean",
    y = "Mean Score",
    x = "Group"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")

# Calculate additional statistics for interpretation
stats_df <- base::data.frame(
  statistic = base::c(
    "Overall Mean",
    "Male Mean",
    "Female Mean",
    "Difference (F-M)",
    "Cohen's d",
    "Distribution Overlap",
    "Upper Decision Limit",
    "Lower Decision Limit",
    "Significant Difference?"
  ),
  value = base::c(
    base::round(overall_mean, 2),
    male_mean,
    female_mean,
    base::round(female_mean - male_mean, 2),
    base::round((female_mean - male_mean) / pooled_sd, 2),
    base::round(calculate_overlap((female_mean - male_mean) / pooled_sd), 1),
    base::round(upper_decision_limit, 2),
    base::round(lower_decision_limit, 2),
    base::ifelse(any(anom_df$significant), "Yes", "No")
  )
)

# Print statistics
print(stats_df)
```

The key differences in an Analysis of Means approach are:

We're comparing each group's mean to the overall mean (rather than directly comparing groups)
We calculate decision limits based on the overall variance
Groups with means outside these limits are considered significantly different from the average
The visual representation focuses on whether groups differ from the overall mean

This approach is particularly useful when you have more than two groups, as it provides a single comprehensive analysis rather than multiple pairwise comparisons.
